The effects of binarity on planet occurrence rates measured by transit surveys

### 2017-11-14 paper/simplest_effects.tex
paper: improved clarity; corrected eqn error

The single stars have luminosity $L_1$, and the twin binaries have luminosity 
$2L_1$.
Thus the twin binaries are selected out to a distance $\sqrt{2}$ times 
that of single stars.
This is a bad move, because the transit signal for any planet in a 
twin binary will be diluted by a factor of two:
\begin{equation}
\frac{{\rm signal}}{{\rm noise}}
\propto \frac{\mathcal{D} (r/R)^2}{F^{-1/2}}
\propto \mathcal{D} F^{1/2}
\propto \mathcal{D} L_{\rm sys}^{1/2} d^{-1}, \quad ({\rm true})
\end{equation}
where the dilution is $\mathcal{D} \equiv L_{\rm host}/L_{\rm sys}$, for 
$L_{\rm host}$ the the planet host's luminosity.
This means that the actual maximum searchable distance for binaries, $d_{\rm 
det}$, is $1/\sqrt{2}$ times that of single stars.
The situation is illustrated in Fig.~\ref{fig:model_1_volumes}: only one in 
eight selected stars in binaries are truly searchable.

@@

\paragraph{What do the observers ignoring binarity infer?} 
The binarity-ignoring observers assume that all points on the sky with flux 
above some minimum are searchable.
They measure a detected planet rate density, and infer a rate 
density by correcting for the transit probability.
As a function of apparent radius, the apparent rate density is
\begin{equation}
\Gamma_a(r_a) = 
\delta(r_p) Z_0 \frac{N_0}{N_0+N_1}  +
\delta\left(\frac{r_p}{\sqrt{2}}\right) 
(Z_1 p_{{\rm det},1} + Z_2 p_{{\rm det},2}) \frac{N_1}{N_0+N_1},
\label{eq:model_1_apparent_rate_density}
\end{equation}
where $p_{{\rm det},1}$ ($p_{{\rm det},2}$) is the probability that a selected 
primary (secondary) is searchable.
In this example, $p_{{\rm det},1}=p_{{\rm det},2}=1/8$; the detection 
efficiency is the ratio of the searchable to selected volumes.

The true rate density (Eq.~\ref{eq:model_1_rate_density}) and the
apparent rate density (Eq.~\ref{eq:model_1_apparent_rate_density})
differ in that
\begin{enumerate}
\item The total number of selected stars, $N_{\rm tot} = N_0+N_1+N_2$, was 
miscounted.
%
\item The detection efficiency was incorrectly assumed to be $1$ for all 
selected stars. In reality, only one in eight binaries were searchable.
%
\item The inferred radii in binary systems are all $\sqrt{2}$ too small.
\end{enumerate}
By making an assumption about the true number of planets per single, primary, 
and secondary star ($Z_0,Z_1,$ and $Z_2$), we can assess the severity of 
these errors.
Taking $Z_0=Z_1=Z_2$, we plot the resulting occurrence rates over bins 
in planet radius in Fig.~\ref{fig:errcases_model_1}.
Though it is impossible for an observer to isolate any one of the above 
errors without also addressing the others, mathematically it is simply a 
matter of modifying the appropriate terms in 
Eq.~\ref{eq:model_1_apparent_rate_density}.
The resulting rates are also shown in 
Fig.~\ref{fig:errcases_model_1}, and demonstrate how each 
error individually affects the overall inferred rate.

\begin{figure}[!tb]
    \begin{center}
        \includegraphics[width=.6\textwidth]{figures/errcases_rate_density_vs_radius_model_1_brokenx.pdf}
    \end{center}
    \vspace{-0.5cm}
    \caption{
        Inferred planet occurrence rates over $0.01r_\oplus$ bins in planet 
        radius,
        for Model \#1.
        This model has fixed stars, fixed planets, and twin binaries.
        If the true planet radius is $r_p$, all planets 
        detected in binaries will have apparent radii $r_a = r_p/\sqrt{2}$.
        To illustrate the individual biases, we 
        separated them:
        ``only wrong $N_\star$'' means the only error is an incorrectly 
        assumed 
        number of selected stars;
        ``only wrong $Q$'' means the only error is an incorrectly assumed 
        detection efficiency (including both miscalculated transit 
        probabilities 
        and 
        fraction of selected stars that are searchable);
        ``only wrong $r_a$'' means the only error is misinterpretation of 
        planetary radii, due to both transit depth dilution and also wrongly 
        assumed host star radii.
    }
    \label{fig:errcases_model_1}
\end{figure}

@@

Still assuming that the number of planets per single, primary, and 
secondary star are equal ($Z_0=Z_1=Z_2$), we find a rate density correction 
factor at the true planet radius of
\begin{equation}
X_\Gamma(r_p) = (1+\mu)^{-1},
\label{eq:Z_eq_model_1_correction}
\end{equation}
where 
\begin{align}
\mu \equiv \frac{N_1}{N_0} &=
\frac{n_b}{n_s} \left(\frac{d_{\rm sel,b}}{d_{\rm sel,s}}\right)^3 = 
\frac{{\rm BF}}{1-{\rm BF}} (1+\ell)^{3/2}.
\label{eq:mu_definition}
\end{align}
In Eq.~\ref{eq:mu_definition}, $n_b$ and $n_s$ are the number density of 
binaries and singles in a volume limited sample, and the maximum selected 
distance for binary and single systems are $d_{\rm sel,b}$ and $d_{\rm sel,s}$.
The light ratio $\ell$ is defined relative to the primary, so that
$L_{\rm sys} = L_1(1+\ell)$.
The binary fraction in a volume limited sample is ``{\rm BF}''.
For instance, Raghavan et al., (2010) found a multiplicity fraction\footnote{
    The binary fraction is the fraction of systems in a volume-limited sample 
    that are binary. It is equivalent to the multiplicity fraction if there 
    are no triple, quadruple, or higher order multiples. In that case, ${\rm 
    BF} = n_b / (n_s+n_b)$.
}
for primaries with masses $0.7-1.3M_\odot$ of 0.44. 
``Twin binaries'' are perhaps 10-20\% of the binaries in a volume-limited 
sample, depending on how ``twin'' is defined.
Thus a representative twin binary fraction to use in
Eq.~\ref{eq:Z_eq_model_1_correction} would be ${\rm BF}\approx 
0.05$-$0.1$. This yields a rate density correction $X_\Gamma(r_p)$ of 0.79 if 
${\rm BF}=0.1$, and 0.88 if ${\rm BF}=0.05$.
Rephrasing the result, the apparent rate would be half the true rate 
if the twin binary fraction were $(2^{3/2}+1)^{-1}\approx 0.26$.
Fortunately, in most contexts it is not that high.

An alternative assumption is that secondaries do not host planets. In this 
case, $Z_0=Z_1$, and $Z_2=0$. The correction to the rate density at the true 
planet radius becomes
\begin{equation}
X_\Gamma(r_p) = (1+2\mu)/(1+\mu)^2.
\end{equation}
This evaluates to 0.95 if ${\rm BF}=0.1$, and 0.99 if ${\rm BF}=0.05$.
While it is hard to justify the assumption that secondaries are planet-less, 
it is worth noting that $\Gamma_a/\Gamma$ is sensitive to the relative number 
of planets per single, primary, and secondary.

>>>>

The single stars have luminosity $L_1$, and the twin binaries have luminosity 
$2L_1$.
Thus the twin binaries are selected out to a distance $\sqrt{2}$ times 
that of single stars.
This is a bad move, because the transit signal for any planet in a 
twin binary will be diluted by a factor of two:
\begin{equation}
\frac{{\rm signal}}{{\rm noise}}
\propto \frac{\mathcal{D} (r/R)^2}{F^{-1/2}}
\propto \mathcal{D} F^{1/2}
\propto \mathcal{D} L_{\rm sys}^{1/2} d^{-1}, \quad ({\rm true})
\end{equation}
where the dilution is $\mathcal{D} \equiv L_{\rm host}/L_{\rm sys}$, for 
$L_{\rm host}$ the the planet host's luminosity.
This means that the true maximum searchable distance for binaries, $d_{\rm 
det}$, is $1/\sqrt{2}$ times that of single stars.
The situation is illustrated in Fig.~\ref{fig:model_1_volumes}: only one in 
eight selected stars in binaries are truly searchable.

@@

\paragraph{What do the observers ignoring binarity infer?} 
The binarity-ignoring observers assume that all points on the sky with flux 
above some minimum are searchable.
They correct their assumed number of searchable stars for the transit 
probability.
As a function of apparent radius, they then report an apparent rate density of
\begin{equation}
\Gamma_a(r_a) = 
\delta(r_p) Z_0 \frac{N_0}{N_0+N_1}  +
\delta\left(\frac{r_p}{\sqrt{2}}\right) 
(Z_1 p_{{\rm det},1} + Z_2 p_{{\rm det},2}) \frac{N_1}{N_0+N_1},
\label{eq:model_1_apparent_rate_density}
\end{equation}
where $p_{{\rm det},1}$ ($p_{{\rm det},2}$) is the probability that a selected 
primary (secondary) is searchable.
In this example, $p_{{\rm det},1}=p_{{\rm det},2}=1/8$; the detection 
efficiency is the ratio of the searchable to selected volumes.

The true rate density (Eq.~\ref{eq:model_1_rate_density}) and the
apparent rate density (Eq.~\ref{eq:model_1_apparent_rate_density})
differ in that
\begin{enumerate}
\item The total number of selected stars, $N_{\rm tot} = N_0+N_1+N_2$, was 
miscounted.
%
\item The detection efficiency was incorrectly assumed to be $1$ for all 
selected stars. In reality, only one in eight binaries were searchable.
%
\item The inferred radii in binary systems are all $\sqrt{2}$ too small.
\end{enumerate}

To assess the severity of these errors, we need to make assumptions about the 
stellar and planetary populations.
The important quantity for the stellar population is $N_1/N_0$, the ratio of 
selected primaries (or binaries) to singles. We can define
\begin{align}
\mu \equiv \frac{N_1}{N_0} &=
\frac{n_b}{n_s} \left(\frac{d_{\rm sel,b}}{d_{\rm sel,s}}\right)^3 = 
\frac{{\rm BF}}{1-{\rm BF}} (1+\ell)^{3/2},
\label{eq:mu_definition}
\end{align}
where $n_b$ and $n_s$ are the number density of binaries and singles in a 
volume limited sample, and the maximum selected distance for binary and single 
systems are $d_{\rm sel,b}$ and $d_{\rm sel,s}$.
The light ratio $\ell$ is defined relative to the primary, so that
$L_{\rm sys} = L_1(1+\ell)$.
The binary fraction in a volume limited sample is ``{\rm BF}''.
For instance, Raghavan et al., (2010) found a multiplicity fraction\footnote{
    The binary fraction is the fraction of systems in a volume-limited sample 
    that are binary. It is equivalent to the multiplicity fraction if there 
    are no triple, quadruple, or higher order multiples. In that case, ${\rm 
        BF} = n_b / (n_s+n_b)$.
} of 0.44 for primaries with masses from $0.7M_\odot$ to $1.3M_\odot$. 
``Twin binaries'' are perhaps 10-20\% of the binaries in a volume-limited 
sample, depending on how ``twin'' is defined.
Thus a representative twin binary fraction for this model is ${\rm BF}\approx 
0.05$-$0.1$.

Further, we need to assume knowledge of the true number of planets per 
single, primary, and secondary star ($Z_0,Z_1,$ and $Z_2$).
Taking $Z_0=Z_1=Z_2=1/2$, we plot the resulting occurrence rates as a function 
of planet radius in Fig.~\ref{fig:errcases_model_1}.
Though it is impossible for an observer to isolate any one of binarity's 
biases without also addressing the others, mathematically it is simply a 
matter of modifying the appropriate terms in 
Eq.~\ref{eq:model_1_apparent_rate_density}.
The resulting rates are also shown in 
Fig.~\ref{fig:errcases_model_1}, and demonstrate how each 
error individually affects the overall inferred rate.

\begin{figure}[!tb]
    \begin{center}
        \includegraphics[width=.6\textwidth]{figures/errcases_rate_density_vs_radius_model_1_brokenx.pdf}
    \end{center}
    \vspace{-0.5cm}
    \caption{
        Inferred planet occurrence rates over $0.01r_\oplus$ bins in planet 
        radius,
        for Model \#1.
        This model has fixed stars, fixed planets, and twin binaries.
        If the true planet radius is $r_p$, all planets 
        detected in binaries will have apparent radii $r_a = r_p/\sqrt{2}$.
        To illustrate the individual biases, we 
        separated them.
        On this and future plots,
        ``only wrong $N_\star$'' means the only error is an incorrectly 
        assumed 
        number of selected stars;
        ``only wrong $Q$'' means the only error is an incorrectly assumed 
        completeness (including both miscalculated $p_{\rm tra}$ 
        and fraction of selected stars that are searchable, $p_{\rm det}$);
        ``only wrong $r_a$'' means the only error is in miscalculated        
        planetary radii, due to both transit depth dilution and also wrongly 
        assumed host star radii.
    }
    \label{fig:errcases_model_1}
\end{figure}

@@

If we continue assuming that the number of planets per single, primary, and 
secondary star are equal ($Z_0=Z_1=Z_2$), we find a rate density correction 
factor at the true planet radius of
\begin{equation}
X_\Gamma(r_p) = \frac{1}{1+\mu},
\label{eq:Z_eq_model_1_correction}
\end{equation}
This yields a correction of 0.79 if ${\rm BF}=0.1$, and 0.88 if ${\rm 
BF}=0.05$.
Rephrasing the result, if the twin binary fraction were 
$(2^{3/2}+1)^{-1}\approx 0.26$, then the apparent rate would be half the true 
rate.
Fortunately, in most contexts the twin binary fraction is not that high.

An alternative assumption is that secondaries do not host planets. In that 
case, $Z_0=Z_1$, and $Z_2=0$. The correction to the rate density at the true 
planet radius becomes
\begin{equation}
X_\Gamma(r_p) = \frac{1+2\mu}{(1+\mu)^2}.
\end{equation}
This evaluates to 0.95 if ${\rm BF}=0.1$, and 0.99 if ${\rm BF}=0.05$.
While it is hard to justify the assumption that secondaries are planet-less, 
it is worth noting that $\Gamma_a/\Gamma$ is sensitive to the relative number 
of planets per single, primary, and secondary.
